---
bibliography: references.bib
---

### Load required packages

```{r}
#load packages   
library(terra)  
library(tidyverse) 
library(sf) 
library(PROJ) 
```

### Data sources & import:

1.  Regional land degradation assessment for the Hardeveld Bioregion conducted by Bell et al in 2021 [@bell2021].

2.  National Land Cover (7 class version prepared by SANBI) 2022.

3.  National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI [@rutherford2006; @Dayaram2019].

```{r}
# Run LCchange_7class_veg24.qmd data loading steps prior to running this code

# Add Hardeveld degradation data (prepared in ARCGIS PRO), where 0 = inside AOI but level of degradation < that 80%; where 8 = severely degraded >= 80%). 
hv_path <- list.files(
  path = data_path,
  pattern = "^hv_bell\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

hv <- terra::rast(hv_path)
hv <- resample(hv, lc2022, method = "near") #  match extent

```

## Analysis

*Cross-tabulate land cover time series with vegetation.*

Step 1. Stack the land cover for each time period with the vegetation data.

Step 2. Cross-tabulate each raster stack and add a column denoting the YEAR (from Land Cover) to the output table (in tibble format). Rename each output column as "LC" to allow binding later.\

```{r}
# Cross Tabulate veg condition hardeveld and lc2022
hv_rall  <- as_tibble(crosstab(c(lc2022, vegrast, hv)))   
write.csv(hv_rall, file = "outputs/hv_rall.csv")

# Import cross tab output to avoid running spatial analysis 
hv_rall <- read.csv("outputs/hv_rall.csv")

# Calculate extent of each vegtype within the specific AOI 
hv_list <- hv_rall %>%
  filter(hv_bell != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_hv = sum(n)*400/1000000)

# clean up names and add km2
hv_rall  <- hv_rall  %>%    
  mutate(km2 = n*400/1000000) %>% # pixel size is 20m
  rename(LC = nlc2022_7class) %>%
  rename(HV = hv_bell)

options(scipen =999)
hv_sum <- hv_rall %>%
  filter(T_MAPCODE != "NOTVEG") %>%
  mutate(LCcomb = case_when(
                            HV == 8 ~ "deg80", 
                            LC == 1 ~ "nat", 
                            LC > 1 ~ "notnat", 
                            TRUE ~ "unknown")) %>%
  group_by(T_MAPCODE, LCcomb) %>%
  summarise(km2tot = sum(km2)) %>%
  pivot_wider(names_from = LCcomb, values_from = km2tot) %>%
    mutate(prcdeg_hv = 100*round(deg80/(nat+deg80),2)) %>% # cond of remnants (D3)
    mutate(D3_hv = case_when(prcdeg_hv >= 90 ~ "EN",
                   prcdeg_hv >= 70 ~ "VU",
                   TRUE ~ "LC")) %>%
    left_join(hv_list, by = "T_MAPCODE") %>%
   mutate(ext_hv = ext_hv/(nat + notnat + deg80))

write.csv(hv_sum, file = "outputs/hv_sum.csv")
  
```
