---
bibliography: references.bib
---

### Load required packages

```{r}
#load packages   
library(terra)  
library(tidyverse) 
library(sf) 
library(PROJ) 
```

### Data sources & import:

1.  Regional land degradation assessment for the Little Karoo conducted by Kirsten et al in 2023 [@kirsten2023].

2.  National Land Cover (7 class version prepared by SANBI) 2022.

3.  National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI [@rutherford2006; @Dayaram2019].

```{r}
# Run LCchange_7class_veg24.qmd data loading steps prior to running this code

# Add Little Karoo degradation data (prepared in ARCGIS PRO), where 0 = inside AOI but level of degradation < that 80%; where 8 = severely degraded >= 80%). 

lkkir_path <- list.files(
  path = data_path,
  pattern = "^lk_kirsten\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

lkkir <- terra::rast(lkkir_path)
lkkir <- resample(lkkir, lc2022, method = "near") #  match extent
```

## Analysis

*Cross-tabulate land cover time series with vegetation.*

Step 1. Stack the land cover for each time period with the vegetation data.

Step 2. Cross-tabulate each raster stack and add a column denoting the YEAR (from Land Cover) to the output table (in tibble format). Rename each output column as "LC" to allow binding later.\

```{r}
# Analysis #Step 1-2.   # get LC  at 1990 time period 
lkk_rall  <- as_tibble(crosstab(c(lc2022, vegrast, lkkir)))   
write.csv(lkk_rall, file = "outputs/lkk_rall.csv")

# Import cross tab output to avoid running spatial analysis 
#lkk_rall <- read.csv("outputs/lkk_rall.csv")

# Calculate extent of each vegtype within the specific AOI 
lkk_list <- lkk_rall %>%
  filter(lk_kirsten != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_lkk = sum(n)*400/1000000)

# clean up names and add km2
lkk_rall  <- lkk_rall  %>%    
  mutate(km2 = n*400/1000000) %>% # pixel size is 20m
  rename(LC = nlc2022_7class) %>%
  rename(LKK = lk_kirsten)

options(scipen =999)
lkk_sum <- lkk_rall %>%
  filter(T_MAPCODE != "NOTVEG") %>%
  mutate(LCcomb = case_when(
                            LKK == 8 ~ "deg80", 
                            LC == 1 ~ "nat", 
                            LC > 1 ~ "notnat", 
                            TRUE ~ "unknown")) %>%
  group_by(T_MAPCODE, LCcomb) %>%
  summarise(km2tot = sum(km2)) %>%
  pivot_wider(names_from = LCcomb, values_from = km2tot) %>%
    mutate(prcdeg_lkk = 100*round(deg80/(nat+deg80),2)) %>% # cond of remnants (D3)
    mutate(D3_lkk = case_when(prcdeg_lkk >= 90 ~ "EN",
                   prcdeg_lkk >= 70 ~ "VU",
                   TRUE ~ "LC")) %>%
    left_join(lkk_list, by = "T_MAPCODE") %>%
   mutate(ext_lkk = ext_lkk/(nat + notnat + deg80))

write.csv(lkk_sum, file = "outputs/lkk_sum.csv")
  
```
