---
bibliography: references.bib
---

### Load required packages

```{r}
#load packages   
library(terra)  
library(tidyverse) 
library(sf) 
library(PROJ) 
```

### Data sources & import:

1.  Little Karoo degradation map developed in 2009 [@thompson2009]. The severely degraded class from this data is estimated to be equivalent to 80 severity on RLE Criterion D - factors include overgrazing / browsing resulting in severe reduction in shrub canopy cover, changes in species composition (loss of palatable species usually dominant), increase in bare ground fraction (with soil loss). Most impacts occurred more than 50 years ago, and in this arid region the shrubland does not recover naturally over time - rather it enters an alternative stable state - bare ground with annual grass and herbs following rainfall events - limited perennial cover).

2.  National Land Cover (7 class version prepared by SANBI) 2022.

3.  National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI [@rutherford2006; @Dayaram2019].

```{r}
# Run LCchange_7class_veg24.qmd data loading steps prior to running this code.

# Add Little Karoo degradation data 2009 (prepared in ARCGIS PRO), where 0 = inside AOI but level of degradation < that 80%; where 8 = severely degraded >= 80%). 

lk_path <- list.files(
  path = data_path,
  pattern = "^littlekaroo_deg4\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

lk <- terra::rast(lk_path)
lk <- resample(lk, lc2022, method = "near") #  match extent
```

## Analysis

*Cross-tabulate land cover time series with vegetation.*

Step 1. Stack the land cover for each time period with the vegetation data.

Step 2. Cross-tabulate each raster stack and add a column denoting the YEAR (from Land Cover) to the output table (in tibble format). Rename each output column as "LC" to allow binding later.\

```{r}
# Crosstab little Karoo, nlc2022, and veg
# Analysis #Step 1-2.   # get LC  at 2022 time period 
lk_rall  <- as_tibble(crosstab(c(lc2022, vegrast, lk)))   
write.csv(lk_rall, file = "outputs/lk_rall.csv")

# Import cross tab output to avoid running spatial analysis 
#lk_rall <- read.csv("outputs/lk_rall.csv")

# Calculate extent of each vegtype within the specific AOI 
lk_list <- lk_rall %>%
  filter(littlekaroo_deg4 != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_lk = sum(n)*400/1000000)

# clean up names and add km2
lk_rall  <- lk_rall  %>%    
  mutate(km2 = n*400/1000000) %>% # pixel size is 20m
  rename(LC = nlc2022_7class) %>%
  rename(LK = littlekaroo_deg4)

options(scipen =999)
lk_sum <- lk_rall %>%
  filter(T_MAPCODE != "NOTVEG") %>%
  mutate(LCcomb = case_when(
                            LK == 8 ~ "deg80", 
                            LC == 1 ~ "nat", 
                            LC > 1 ~ "notnat", 
                            TRUE ~ "unknown")) %>%
  group_by(T_MAPCODE, LCcomb) %>%
  summarise(km2tot = sum(km2)) %>%
  pivot_wider(names_from = LCcomb, values_from = km2tot) %>%
    mutate(prcdeg_lk = 100*round(deg80/(nat+deg80),2)) %>% # cond of remnants (D3)
    mutate(D3_lk = case_when(prcdeg_lk >= 90 ~ "EN",
                   prcdeg_lk >= 70 ~ "VU",
                   TRUE ~ "LC")) %>%
    left_join(lk_list, by = "T_MAPCODE") %>%
   mutate(ext_lk = ext_lk/(nat + notnat + deg80))

write.csv(lk_sum, file = "outputs/lk_sum.csv")
  
```
