---
title: "Western Cape (2020), KwaZulu-Natal (2017)  & Mpumalanga (2017) Land Cover"
format: html
execute:
  freeze: auto
---

### Load required packages

```{r}
#load packages   
library(terra)  
library(tidyverse) 
library(sf) 
library(PROJ) 
```

### Data sources & import:

1.  Western Cape Provincial Land Cover (Cape Nature, 2022). Data supplied in ESRI GRID format, converted to a TIFF in ARCGIS PRO, 10 m data (UTM34S) resampled to match 20m and Albers Equal Area national land cover grid. Reclassified to match the NLC classes.

2.  KwaZulu-Natal Provincial Land Cover (Ezemvelo KZN Wildlife, 2017). Data supplied in ESRI gdb format, converted to a TIFF in ARCGIS PRO, 20 m data (UTM35S) resampled to match 20m and Albers Equal Area national land cover grid. Reclassified to match the NLC classes.

3.  Mpumalanga Provincial Land Cover (MPTA, 2017). Data supplied in ESRI gdb format, converted to a TIFF in ARCGIS PRO, 20 m data (UTM35S) resampled to match 20m and Albers Equal Area national land cover grid. Reclassified to match the NLC classes.

4.  National Land Cover (7 class version prepared by SANBI) 2022.

5.  National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI [@rutherford2006; @Dayaram2019].

    ```{mermaid}
    flowchart LR
    A(Western Cape data) -- reclass and resample ARCPRO--> B(wc_lc2.tif)
    C(KwaZulu-Natal data) --reclass and resample ARCPRO--> D(kzn_lc2.tif)
    E(Mpumalanga data) --reclass and resample ARCPRO--> F(mpl_lc3.tif)
    B --load to R terra--> L
    D --load to R terra--> L
    I[NLC 2022 DFFE]--reclass in ARCPRO--> J(nlc2022_7class.tif) --load to R terra--> K(Resample to 30m national grid)
    K--> L(Cross tabulate)
    M[National Vegetation Map 2024] --load and make raster --> L
    L--> N(Summarise)
    ```

```{r}

# Run LCchange_7class_veg24.qmd data loading steps prior to running this code

# Add WC, KZN and MPL LC data  (prepared in ARCGIS PRO), where 0 = inside AOI but no value assigned, 2 = old fields, 4 = lumped crop/built/mine/plantation/dams).
 
wc_path <- list.files(
  path = data_path,
  pattern = "^wc_lc3\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

wc <- terra::rast(wc_path)
wc <- resample(wc, lc2022, method = "near") #  match extent
########################
kzn <- list.files(
  path = data_path,
  pattern = "^kzn_lc2\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

kzn <- terra::rast(kzn_path)
kzn <- resample(kzn, lc2022, method = "near") #  match extent
#########################
mp_path <- list.files(
  path = data_path,
  pattern = "^mpl_lc3\\.tif$",
  recursive = TRUE,
  full.names = TRUE,
  ignore.case = TRUE
)

mp <- terra::rast(mp_path)
mp <- resample(mp, lc2022, method = "near") #  match extent
```

## Analysis

*Cross-tabulate land cover vegetation.*

Step 1. Stack the land cover for region with national land cover and national vegetation data.

Step 2. Cross-tabulate raster stack and add a column region to the output table (in tibble format).

```{r}
# Analysis #Step 1-2.   # get LC  at 1990 time period 
alt_rall  <- as_tibble(crosstab(c(lc2022, vegrast, wc, mp, kzn)))   
write.csv(alt_rall, file = "outputs/alt_rall.csv")

# Import cross tab output to avoid running spatial analysis 
alt_rall <- read.csv("C:/Rdata/LCC_terr/outputs/alt_rall.csv")

# clean up names and add km2
alt_rall  <- alt_rall  %>%    
  mutate(km2 = n*400/1000000) %>%   # pixel size is 40m
  rename(LC = nlc2022_7class) %>%
  rename(WC = wc_lc3) %>%
  rename(KZN = kzn_lc2) %>%
  rename(MPL = mpl_lc3) 

# Calculate extent of each vegtype within the specific AOI 
wc_list <- alt_rall %>%
  filter(WC != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_wc = sum(km2))

kzn_list <- alt_rall %>%
  filter(KZN != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_kzn = sum(km2))
  
mpl_list <- alt_rall %>%
  filter(MPL != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_mpl = sum(km2))

# Summarise data and assess Criterion A3
alt_sum <- alt_rall %>%
  filter(T_MAPCODE != "NOTVEG") %>%
  mutate(LCcomb = case_when(
    MPL == 99 & WC == 99 & KZN == 99 & LC == 1 ~ "nat", 
    MPL == 99 & WC == 99 & KZN == 99 & LC > 1 ~ "notnat", 
    WC > 1 & WC < 99 ~ "notnat",
    WC == 1 ~ "nat", 
    KZN == 4 | KZN == 2 ~ "notnat", 
    KZN == 0 ~ "nat", 
    MPL == 4 ~ "notnat", 
    MPL == 0 ~ "nat", 
    TRUE ~ "unknown"  # Catch-all for unexpected values
  )) %>% 
  group_by(T_MAPCODE, LCcomb) %>%
  summarise(km2tot = sum(km2)) %>%
  pivot_wider(names_from = LCcomb, values_from = km2tot) %>%
    mutate(prclost_alt = 100*round(notnat/(nat+notnat),2)) %>% # ext of remnants (A3)
    mutate(A3_alt = case_when(prclost_alt >= 90 ~ "CR",
                   prclost_alt >= 70 ~ "EN",
                   prclost_alt >= 50 ~ "VU",
                  TRUE ~ "LC")) %>% 
   left_join(wc_list, by = "T_MAPCODE") %>%
   left_join(kzn_list, by = "T_MAPCODE") %>%
   left_join(mpl_list, by = "T_MAPCODE") %>%
   mutate(ext_wc = ext_wc/(nat + notnat)) %>%
   mutate(ext_kzn = ext_kzn/(nat + notnat)) %>%
   mutate(ext_mpl = ext_mpl/(nat + notnat)) 

write.csv(alt_sum, file = "outputs/alt_sum.csv")
  
```
